\section{Introduction}
\label{sec:intro}

% P1 — Scene: application-level framing
Sparse convolution is the computational backbone of 3D point-cloud
perception.
LiDAR-based object detection~\cite{centerpoint,second}, semantic
segmentation~\cite{semantickitti}, and scene reconstruction all build on
stacks of sparse convolution layers to process inherently sparse voxelized
representations of 3D sensor data.
In production models such as CenterPoint~\cite{centerpoint} and
SECOND~\cite{second}, sparse convolution accounts for
\textbf{TODO}\% of end-to-end inference time, making its runtime
efficiency a first-order concern.
These models operate at extreme sparsity: across a typical detection
network on KITTI, the average per-layer active ratio is below
\textbf{TODO}\%, and even the densest layers rarely exceed
\textbf{TODO}\%.

% P2 — What sparse convolution computes and how it differs from sparse MM
Unlike the sparse matrix multiplications common in large language models,
where sparsity appears in the channel (feature) dimension and the
computation pattern is structurally regular, 3D sparse convolution faces a
fundamentally different challenge.
A sparse convolution with kernel size $k$ (typically 3; we use symmetric
$k^3$ kernels throughout) examines $k^3$ neighbor offsets around each
active output voxel.
The feature channels are always dense; what is sparse is the set of
\emph{spatially active voxels} and, consequently, the set of active
neighbors at each offset.
At each offset, the runtime checks whether the corresponding input
neighbor is also active; if so, the input features at that neighbor are
multiplied against the offset's weight slice and accumulated into the
output.
The set of active neighbors varies per voxel and per offset, producing an
irregular, data-dependent computation graph that changes with every input.
The central runtime challenge is how to organize this spatially irregular
work into efficient GPU execution.

% P3 — Existing approaches: what they get and what they lose
Existing systems commit to one side of a fundamental tradeoff between
\emph{computation skipping} and \emph{execution efficiency}.

The \emph{output-stationary} approach (implicit GEMM; SpConv~v2~\cite{spconv2},
TorchSparse++~\cite{torchsparse++}) assigns each thread block a set of
output voxels and iterates over all $k^3$ offsets, treating every neighbor
as present.
The full neighborhood is concatenated into one matrix and a single fused
dense multiply is executed across all offsets simultaneously.
This yields strong output locality (all partial sums for one output
accumulate in-block, no atomic writes needed), good weight reuse (all
outputs in a block share the same weight slice at each offset step), and
high hardware utilization.
But it cannot skip empty neighbors: missing neighbors contribute
zero-valued rows that still consume compute and memory bandwidth.
TorchSparse++ mitigates this via bitmask sorting (warp-level K-tile skip),
but the skip granularity is limited by the output-stationary framework.
Profiling a CenterPoint backbone on KITTI reveals that \textbf{TODO}\%
of the fused multiply-accumulate operations involve zero-valued inputs.
The GPU looks busy, but useful utilization is low.

The \emph{weight-stationary} approach (per-offset gather-GEMM-scatter;
MinkowskiEngine~\cite{minkowski}) takes the opposite route: it processes
offsets one by one, collecting only the active pairs at each offset and
running a separate matrix multiply for each.
This achieves exact work (computation proportional to active pairs) and
good weight locality within each offset's multiply.
However, different offsets have vastly different numbers of active pairs,
producing severe load imbalance.
This imbalance is fundamental (it comes from the spatial structure of the
data); MinkowskiEngine's 27 separate kernel launches make it worse, but
even a fused single-kernel per-offset approach still suffers from the
uneven pair distribution.
The same output voxel also receives partial sums from multiple offsets
across separate multiplies, requiring scatter and accumulation.

At the low densities that dominate real workloads, both approaches result
in poor GPU utilization on useful work.
Output-stationary keeps the SM busy but wastes most compute on empty
neighbors.
Weight-stationary does only useful compute but cannot spread it evenly
across the GPU.
The total amount of useful work (active neighbor pairs) is small; the
first-order priority is to maximize the fraction of GPU capacity spent on
that useful work.

% P4 — Our approach: disassemble for utilization
\sysname starts from the weight-stationary side (per-offset exact work)
and disassembles it for utilization.
We precompute all active \texttt{(input, output)} neighbor pairs across
every offset, group them by offset, and flatten the groups into a
contiguous stream of dense GEMM tiles: a worklist.
Instead of each offset owning a separate multiply with its own resource
allocation, all offsets' tiles are pooled into one flat stream and
distributed evenly by a single kernel.
This design has three properties:
%
\begin{itemize}
  \item \textbf{Exact work.}
        Offsets with no active pairs produce zero tiles and are skipped
        entirely; computation is strictly proportional to active neighbor
        pairs.
  \item \textbf{Global load balance.}
        The flat tile stream is divided evenly across thread blocks.
        No offset is too small to fill a block or too large and starves
        others.
  \item \textbf{Dense inner compute.}
        Each tile is a standard dense matrix multiply, fully GPU-friendly
        and Tensor Core compatible.
\end{itemize}

% P5 — What the worklist loses: locality
The worklist maximizes utilization, but it disrupts both directions of
locality that the original approaches had.
Output-stationary's ``same output, all offsets in one block'' is gone:
the same output voxel's partial sums are now spread across tiles from
different offsets, requiring atomic additions.
Weight-stationary's ``same offset, contiguous execution'' is partially
preserved (tiles are grouped by offset within the worklist), but a naive
stride-loop assignment can interleave tiles from different offsets across
blocks, breaking weight reuse.
More broadly, dense and implicit-GEMM convolution exploit the fact that
spatially adjacent outputs share overlapping input neighborhoods; the
worklist groups by offset instead, so cross-offset spatial relationships
are lost.

The challenge is how to recover enough locality \emph{without} giving up
exact work or breaking global load balance.

% P6 — Persistent threads: the architectural bridge
The worklist solves \emph{what} to compute; recovering locality requires
control over \emph{how} tiles are scheduled onto hardware.
Under ordinary grid dispatch, each thread block processes one tile and
exits.
The hardware scheduler decides which block runs on which SM next, so
two logically adjacent tiles may land on different SMs, register state
is lost between tiles, and L1/shared-memory contents are overwritten by
intervening blocks.
Any logical ordering of the worklist (e.g., offset-major) would have no
guaranteed physical effect on data reuse.

\sysname uses persistent threads: each thread block stays resident on its
SM and processes tile after tile via a stride loop.
This turns logical tile ordering into physical locality: the block's
register file survives across tile boundaries, L1 cache and shared memory
state carry over, and the tile processing order is exactly the worklist
order.

% P7 — Locality recovery: reassemble
Building on persistent threads, \sysname recovers locality with two
low-complexity dataflow choices rather than a heavy scheduling framework:
%
\begin{itemize}
  \item \textbf{Offset-major contiguous tile assignment.}
        Tiles in the worklist are ordered so that consecutive iterations of
        the stride loop share the same kernel offset.
        This improves L2 cache locality for input features and enables the
        weight reuse described next, while preserving the load-balance
        property of the global stride loop.
  \item \textbf{Cross-tile register reuse.}
        When consecutive tiles share the same offset, the weight fragment
        already in the persistent block's registers carries over directly.
        L2 cache is insufficient for this: a single tile's K-loop touches
        enough working data (indirect input gathers, partial outputs,
        weight BK-slices) to evict previously cached weights.
        Register persistence bypasses this eviction entirely.
        Input-feature fragments can also be reused across output-channel
        (BN) tiles within the same offset, which is especially valuable
        because input loads are indirect gathers.
        As a side effect, keeping data in registers reduces L2 cache
        pressure, freeing capacity for the remaining irregular accesses
        that benefit most from caching.
\end{itemize}
%
The reassembled runtime not only recovers the locality that the worklist
disrupted, but goes further: cross-tile input-feature reuse across
output-channel tiles is something neither the original output-stationary
nor weight-stationary approach provided.

% P8 — Preview table
Table~\ref{tab:intro-perf} previews the operator-level impact of this
execution model.
On an RTX~2080~Ti with FP32 precision, \sysname outperforms all existing
sparse convolution systems across the entire active-ratio range tested.
At 1\% active ratio (typical for LiDAR detection layers), \sysname is
$5.5\times$ faster than the fastest existing system.
The advantage holds up to 30\%, where \sysname still maintains a
$1.8\times$ margin.
End-to-end evaluation on detection and segmentation models is presented
in \S\ref{sec:eval}.

\begin{table}[t]
\centering
\caption{Operator-level kernel latency (ms) for SubM $3{\times}3{\times}3$
convolution, $C_\text{in}{=}C_\text{out}{=}64$, spatial
$128{\times}128{\times}8$, FP32, RTX~2080~Ti.
Speedup is over the fastest existing system at each ratio.
End-to-end model evaluation with real datasets is in
\S\ref{sec:eval} (\emph{in progress}).}
\label{tab:intro-perf}
\small
\begin{tabular}{@{}r rrrr r@{}}
\toprule
Active & \multirow{2}{*}{\sysname} & SpConv & TorchSparse++ & Minkowski- & \multirow{2}{*}{Speedup} \\
ratio  &          & v2     &               & Engine     &  \\
\midrule
0.1\% & \textbf{0.012} & 0.421 & 0.240 & 0.230 & $19.2\times$ \\
0.5\% & \textbf{0.045} & 0.374 & 0.240 & 0.323 & $5.3\times$ \\
1\%   & \textbf{0.045} & 0.403 & 0.246 & 0.379 & $5.5\times$ \\
5\%   & \textbf{0.026} & 0.436 & 0.242 & 0.386 & $9.3\times$ \\
10\%  & \textbf{0.067} & 0.551 & 0.377 & 0.445 & $5.6\times$ \\
20\%  & \textbf{0.240} & 1.251 & 0.685 & 0.908 & $2.9\times$ \\
30\%  & \textbf{0.531} & 1.544 & 0.945 & 1.703 & $1.8\times$ \\
\bottomrule
\end{tabular}
\end{table}

\medskip
\noindent This paper makes the following contributions:
\begin{enumerate}
\item A \textbf{worklist-driven execution model} for sparse convolution
      that achieves computation proportional to active neighbor pairs
      within a single persistent-thread kernel, unifying exact work and
      global load balance.

\item \textbf{Persistent-thread locality recovery} through offset-major
      tile assignment and cross-tile register reuse of both weight and
      input-feature fragments, recovering and exceeding the data reuse of
      existing approaches without heavy scheduling.

\item An end-to-end evaluation on 3D object detection and semantic
      segmentation models showing \textbf{TODO}$\times$ speedup over
      the fastest existing sparse convolution runtime, with numerically
      identical outputs.
\end{enumerate}
